Analysis of the inference of ratings and rankings on Higher Order Networks with complex topologies

### 2025-09-06 draft/draft.tex
progress

A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $f_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
$$
\sum_{i<j} \bigl(f_{ij}-(f_i-f_j)\bigr)^2.
$$
This yields a well-posed optimization problem that admits an elegant reformulation in the language of cochains over simplicial complexes.
The resulting framework for the inference of ratings and rankings is known as {\em HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.

@@

\subsection{Discrete calculus on simplicial complexes}

To define HodgeRank, an abstract simplicial complex has to be defined.
For this, let the vertex set be $V=\{0,1,\dots,n\}$, representing the teams, and let the edge set $E$ consist of the unordered pairs $\{i,j\}$ for which a pairwise comparison score $\psi_{ij}$ is available.
Then, the associated abstract simplicial complex is
$$
K = K_0 \cup K_1 \cup K_2,
$$
where
\begin{eqnarray}
K_0 &=& \{\{i\} : i \in V\}, \notag \\  
K_1 &=& \{\{i,j\} : \{i,j\} \in E\}, \notag \\
K_2 &=& \{\{i,j,k\} : \{i,j\},\{j,k\},\{k,i\} \in E\} \notag
\end{eqnarray}
are the sets of 0, 1 and 2 simplices, respectively.
Thus, each team corresponds to a 0-simplex, each comparison to a 1-simplex, and each closed triple of comparisons to a 2-simplex.
By convention, $\{i,j\}=\{j,i\}$ denotes the same 1-simplex, and any permutation of $\{i,j,k\}$ denotes the same 2-simplex.

Next, the simplicial complex is equipped with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.

The vector space of chains $C$ is equipped with the trivial inner product
$$
e_r.e_s = \delta_{rs}
$$
for each pair $r,s \in K$.
Using Einstein's implicit summation convention, an arbitrary chain $c$ in $C$ can be linearly decomposed in the canonical basis as $c=c^se_s$ where the $c^s$ are coefficients in $\RR$.
The vector space of chains admits the following decomposition in direct sums
$$
C = C_0\oplus C_1\oplus C_2
$$
where $C_k$ is the subspace of $C$ spanned by the basis of $k$-chains, i.e. by $\{e_s:s\in K_k\}$.

The covectors $f$ in the dual space $C^*$ of $C$ are called cochains.
The dual basis is denoted by $\{e^s:s\in K\}$ and satisfies
$$
e^r(e_s) = \delta^r{}_s
$$
for each $s,r\in K$.
An arbitrary cochain $f$ in $C^*$ can be linearly decomposed in the dual basis as $f=f_se^s$ where the $f_s$ are coefficients in $\RR$.
The vector space of cochains admits the following decomposition in direct sums
$$
C^* = C_0^*\oplus C_1^*\oplus C_2^*
$$
where $C_k^*$ is the subspace of $C^*$ spanned by the dual basis of $k$-cochains, i.e. by $\{e^s:s\in K_k\}$.

@@

Any $k$-cochain $f$ admits the so called Hodge decomposition
$$
f = g\oplus h\oplus s
$$
where the direct sum components $g\in \rng d_{k-1}$, $h\in \krn L_k$ and $s\in \rng d_k^*$ are unique and are called the gradient, the harmonic and the solenoid components, respectively.
As a consequence, there exists non-unique cochains $w\in C_{k-1}^*$ and $u\in C_{k+1}^*$ such that $g=d_{k-1}w$ and $s=d_k^*u$, that can be found solving for
\begin{eqnarray}
d_kf 
&=&
d_k(g+h+s)
\notag
\\
&=&
d_k s
\notag
\\
&=&
d_k d_k^* u
\notag
\\
&=&
L_{k+1}^{\downarrow} u
\notag
\end{eqnarray}
and
\begin{eqnarray}
d_{k-1}^* f 
&=&
d_{k-1}^*(g+h+s)
\notag
\\
&=&
d_{k-1}^* g
\notag
\\
&=&
d_{k-1}^*d_{k-1} w
\notag
\\
&=&
L_{k-1}^{\uparrow} w,
\notag
\end{eqnarray}
while the harmonic component can be found from $h=f-g-s$.

\subsection{Discrete calculus reformulation of the infererence of ratings}

>>>>

A natural approach to inferring rating scores consistent with the pairwise comparisons is to assign ratings $w_i$ (for $i=0,1,\dots,n$) that minimize the sum of squared discrepancies
\begin{equation}
\label{eq:1}
\sum_{i<j} \bigl(f_{ij}-(w_i-w_j)\bigr)^2.
\end{equation}
This yields a well-posed optimization problem that admits an elegant reformulation in the language of cochains over simplicial complexes.
The resulting framework for the inference of ratings and rankings is known as {\em HodgeRank}~\cite{jiang2011statistical, perotti2025analysis}.

@@

\subsection{Discrete exterior calculus on simplicial complexes}

To review the HodgeRank method for the infrence of ratings and rankings, the theory of discrete exterior calculus on simplicial complexes has to be reviewed first.
For this, consider a vertex set $V=\{0,1,\dots,n\}$.
A simplicial complex is a set
$$
K = K_0 \cup K_1 \cup ... \cup K_n
$$
of $k$-simplices
$$
K_k \subseteq \{\{i_0,...,i_k\} : i_0,...,i_k \in V\}
$$
that satisfies the closure condition: if $s\in K$ and $r\subset s$ is non-empty, then $r\in K$.

To define discrete calculus, the simplicial complex is equipped with a vector space $C$ of vectors called chains, whose canonical basis is denoted by $\{e_s:s\in K\}$.

The vector space of chains $C$ is equipped with the trivial inner product
$$
e_r.e_s = \delta_{rs}
$$
for each pair $r,s \in K$.
Using Einstein's convention where repeated indices assume an implicit summation, an arbitrary chain $c$ in $C$ can be linearly decomposed in the canonical basis as $c=c^se_s$ where the $c^s$ are coefficients in $\RR$.
The vector space of chains admits the following decomposition in direct sums
$$
C = C_0\oplus C_1\oplus C_2
$$
where $C_k$ is the subspace of $C$ spanned by the basis of $k$-chains, i.e. by $\{e_s:s\in K_k\}$.

The covectors $f$ in the dual space $C^*$ of $C$ are called cochains.
The dual basis is denoted by $\{e^s:s\in K\}$ and satisfies
$$
e^r(e_s) = \delta^r{}_s
$$
denotes Kronecker's delta for each $s,r\in K$.
An arbitrary cochain $f$ in $C^*$ can be linearly decomposed in the dual basis as $f=f_se^s$ where the $f_s$ are coefficients in $\RR$.
The vector space of cochains admits the following decomposition in direct sums
$$
C^* = C_0^*\oplus C_1^*\oplus C_2^*
$$
where $C_k^*$ is the subspace of $C^*$ spanned by the dual basis of $k$-cochains, i.e. by $\{e^s:s\in K_k\}$.

@@

Any $k$-cochain $f$ admits the so called Hodge decomposition
$$
f = g\oplus h\oplus s
$$
where the direct sum components $g\in \rng d_{k-1}$, $h\in \krn L_k$ and $s\in \rng d_k^*$ are unique and are called the gradient, the harmonic and the solenoid components, respectively.
As a consequence, there exists non-unique cochains $w\in C_{k-1}^*$ and $u\in C_{k+1}^*$ such that $g=d_{k-1}w$ and $s=d_k^*u$, that can be found solving for
\begin{eqnarray}
\label{eq:2}
d_kf 
&=&
d_k(g+h+s)
%\notag
\\
&=&
d_k s
\notag
\\
&=&
d_k d_k^* u
\notag
\\
&=&
L_{k+1}^{\downarrow} u
\notag
\end{eqnarray}
and
\begin{eqnarray}
\label{eq:3}
d_{k-1}^* f 
&=&
d_{k-1}^*(g+h+s)
%\notag
\\
&=&
d_{k-1}^* g
\notag
\\
&=&
d_{k-1}^*d_{k-1} w
\notag
\\
&=&
L_{k-1}^{\uparrow} w,
\notag
\end{eqnarray}
while
 the harmonic component can be found from $h=f-g-s$.

\subsection{HodgeRank}

Expression~\ref{eq:1} can be reformulated in the language of discrete calculuas as
\begin{equation}
\label{eq:4}
|f-d_0(w)|
\end{equation}
if the scalar components $f_{ij}$ of the 1-cochain $f=f_{ij}e^{ij}$ are identified with the pairwise comparison socres and the 0-cochain scalar components $w_i$ are identified with the ratings.
In fact, the vector of ratings minimizing expression~\ref{eq:4} is the solution of Eq.~\ref{eq:3} for the case $k=1$.
For this, the set of 0-simplices $K_0$ is identified with a set $V$ of teams on a league, the set of 1-simplices $K_1$ is identified with the set $E$ of links or pairwise comparison of teams in the league.
Additionally, to compute the 2-cochain $u$ in order to obtain the solenoid component of the Hodge decomposition of $f$ from Eq.~\ref{eq:2}, the set of 2-simplices is identified with the set of connected triples in the set of comparisons; i.e.
$$
K_2
=
\{
\{i,j,k\}
:
\{i,j\}\in E,
\{j,k\}\in E,
\{i,k\}\in E
\}.
$$
